From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-09 Pi+/paper/finite.tex
Add some text and figures in section 4.4

$\UFin$ is a symmetric monoidal groupoid with unit and tensor given by

\begin{definition}
  \begin{align*}
    I           & \defeq F_{0}    \\
    X \otimes Y & \defeq X \sqcup Y
  \end{align*}
\end{definition}

with unitors, associator, and braiding satisfying Mac Lane's triangle, pentagon, and hexagon coherence laws, and
symmetry. It is the free symmetric monoidal groupoid on one generator.

>>>>

We describe the symmetric monoidal structure of the groupoid $\UFin$.

First, we observe a few equivalences.

\begin{proposition}
  For any $n, m : \Nat$,
  \begin{align*}
    \Fin[0]                & \eqv \bot \\
    \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m] \\
  \end{align*}
  and for any types $X, Y, Z$,
  \begin{align*}
    \bot \sqcup X          & \eqv X \\
    X \sqcup \bot          & \eqv X \\
    (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
    X \sqcup Y             & \eqv Y \sqcup Y \\
  \end{align*}
\end{proposition}

We lift these equivalences to $\UFin$ giving it the (additive) symmetric monoidal structure $(I, \oplus)$, with natural
isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that
types in $\UFin$ are $\hSet$s since they're merely equivalent to $\Fin[n]$ for some $n : \Nat$.

\begin{definition}
  \begin{align*}
    I           & \defeq F_{0}    \\
    X \oplus Y & \defeq X \sqcup Y \\
    \lambda_{X} & : I \oplus X \eqv X \\
    \rho_{X} & : X \oplus I \eqv X \\
    \alpha_{X,Y,Z} & : (X \oplus Y) \oplus Z \eqv X \oplus (Y \oplus Z) \\
    \mathcal{B}_{X,Y} & : X \oplus Y \eqv Y \oplus X
  \end{align*}
\end{definition}

These isomorphisms satisfy the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle,
pentagon, and hexagon identities, and the syllepsis of the braiding, upto 2-paths in $\UFin$.

\begin{proposition}
  % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
  \[\begin{tikzcd}
      {(X \oplus I) \oplus Y} && {X \oplus (I \oplus Y)} \\
      {} & {X \oplus Y}
      \arrow["{\alpha_{X,I,Y}}", from=1-1, to=1-3]
      \arrow[""{name=0, anchor=center, inner sep=0}, "{\rho_{X} \oplus 1_{Y}}"', from=1-1, to=2-2]
      \arrow[""{name=1, anchor=center, inner sep=0}, "{1_{X} \oplus \lambda_{Y}}", from=1-3, to=2-2]
      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
    \end{tikzcd}\]
  % https://q.uiver.app/?q=WzAsNSxbMCwxLCIoKFcgXFxvcGx1cyBYKSBcXG9wbHVzIFkpIFxcb3BsdXMgWiJdLFsxLDAsIihXIFxcb3BsdXMgWCkgXFxvcGx1cyAoWSBcXG9wbHVzIFopIl0sWzIsMSwiVyBcXG9wbHVzIChYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSkiXSxbMiwzLCJXIFxcb3BsdXMgKChYIFxcb3BsdXMgWSkgXFxvcGx1cyBaKSJdLFswLDMsIihXIFxcb3BsdXMgKFggXFxvcGx1cyBZKSkgXFxvcGx1cyBaIl0sWzAsMSwiXFxhbHBoYV97VyBcXG9wbHVzIFgsIFksIFp9Il0sWzEsMiwiXFxhbHBoYV97VyxYLFkgXFxvcGx1cyBafSJdLFszLDIsIjFfe1d9IFxcb3BsdXMgXFxhbHBoYV97WCxZLFp9IiwyXSxbMCw0LCJcXGFscGhhX3tXLFgsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzQsMywiXFxhbHBoYV97VyxYIFxcb3BsdXMgWSxafSIsMl0sWzAsMiwiXFxpZCIsMSx7Im9mZnNldCI6NSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
  \[\begin{tikzcd}
      & {(W \oplus X) \oplus (Y \oplus Z)} \\
      {((W \oplus X) \oplus Y) \oplus Z} && {W \oplus (X \oplus (Y \oplus Z))} \\
      \\
      {(W \oplus (X \oplus Y)) \oplus Z} && {W \oplus ((X \oplus Y) \oplus Z)}
      \arrow["{\alpha_{W \oplus X, Y, Z}}", from=2-1, to=1-2]
      \arrow["{\alpha_{W,X,Y \oplus Z}}", from=1-2, to=2-3]
      \arrow["{1_{W} \oplus \alpha_{X,Y,Z}}"', from=4-3, to=2-3]
      \arrow["{\alpha_{W,X,Y} \oplus 1_{Z}}"', from=2-1, to=4-1]
      \arrow["{\alpha_{W,X \oplus Y,Z}}"', from=4-1, to=4-3]
      \arrow["\id"{description}, shift right=5, draw=none, from=2-1, to=2-3]
    \end{tikzcd}\]
  % https://q.uiver.app/?q=WzAsNixbMSwwLCJYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSJdLFswLDEsIihYIFxcb3BsdXMgWSkgXFxvcGx1cyBaIl0sWzAsMiwiKFkgXFxvcGx1cyBYKSBcXG9wbHVzIFoiXSxbMSwzLCJZIFxcb3BsdXMgKFggXFxvcGx1cyBaKSJdLFsyLDIsIlkgXFxvcGx1cyAoWiBcXG9wbHVzIFgpIl0sWzIsMSwiKFkgXFxvcGx1cyBaKSBcXG9wbHVzIFgiXSxbMSwwLCJcXGFscGhhX3tYLFksWn0iXSxbMSwyLCJcXG1hdGhjYWx7Qn1fe1gsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzIsMywiXFxhbHBoYV97WSxYLFp9IiwyXSxbMyw0LCIxX3tZfSBcXG9wbHVzIFxcbWF0aGNhbHtCfV97WCxafSIsMl0sWzUsNCwiXFxhbHBoYV97WSxaLFh9Il0sWzAsNSwiXFxtYXRoY2Fse0J9X3tYLFkgXFxvcGx1cyBafSJdLFs3LDEwLCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
  \[\begin{tikzcd}
      & {X \oplus (Y \oplus Z)} \\
      {(X \oplus Y) \oplus Z} && {(Y \oplus Z) \oplus X} \\
      {(Y \oplus X) \oplus Z} && {Y \oplus (Z \oplus X)} \\
      & {Y \oplus (X \oplus Z)}
      \arrow["{\alpha_{X,Y,Z}}", from=2-1, to=1-2]
      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y} \oplus 1_{Z}}"', from=2-1, to=3-1]
      \arrow["{\alpha_{Y,X,Z}}"', from=3-1, to=4-2]
      \arrow["{1_{Y} \oplus \mathcal{B}_{X,Z}}"', from=4-2, to=3-3]
      \arrow[""{name=1, anchor=center, inner sep=0}, "{\alpha_{Y,Z,X}}", from=2-3, to=3-3]
      \arrow["{\mathcal{B}_{X,Y \oplus Z}}", from=1-2, to=2-3]
      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
    \end{tikzcd}\]
  % https://q.uiver.app/?q=WzAsMyxbMCwwLCJYIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyBZIl0sWzEsMSwiWSBcXG9wbHVzIFgiXSxbMCwxLCIxX3tYIFxcb3BsdXMgWX0iLDAseyJsZXZlbCI6Miwic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcbWF0aGNhbHtCfV97WCxZfSIsMl0sWzIsMSwiXFxtYXRoY2Fse0J9X3tZLFh9IiwyXSxbNCw1LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
  \[\begin{tikzcd}
      {X \oplus Y} && {X \oplus Y} \\
      & {Y \oplus X}
      \arrow["{1_{X \oplus Y}}", Rightarrow, no head, from=1-1, to=1-3]
      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y}}"', from=1-1, to=2-2]
      \arrow[""{name=1, anchor=center, inner sep=0}, "{\mathcal{B}_{Y,X}}"', from=2-2, to=1-3]
      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
    \end{tikzcd}\]
\end{proposition}
